---
title: "Analysis of Matched Case-Control Studies"
author: "Kamarul Imran Musa"
date: "14 February 2016"
output: 
  pdf_document:
    toc: true
---

# Matched Case-Control Studies Require Special Logistic Regression Approach

To analyze case-control data with pairwise matching we use conditional logistic 
regression. Cases are individually matched to 1 (or up to 4) control subjects based on matching criteria. These matching criteria represent the confounders to be controlled. Each case and matched control(s) are analyzed as separate strata. 
For example, if there are 150 cases, then there are 150 strata that need to be fitted. This is accomplished with the Cox proportional hazards regression functions available in the Survival package.

# Motivation

With matched pairs data the form of the logistic model involves the probability, $\phi$, that in matched pair number i, for a given value of the explanatory
variable the member of the pair is a case. Specifically the model is

$$logit(\phi_{i}) = \alpha_{i} + \beta x$$
The odds that a subject with x = 1 is a case equals $exp(\beta)$ times the odds
that a subject with x = 0 is a case.

# Tutorial 1

## Prepare workspace and data

## Set working directory

```{r}
setwd("E:/Epi_Stat_Matters/LectureNotes2015/Clogit-DrPH-Epid-2015-16")

```

Let's get an overview of the data.

```{r}
library(HSAUR2)
head(backpain)
```

Describe data

```{r}
library(psych)
describe(backpain)
```

Perform survival::clogit

```{r}
library(survival)
backpain_glm <- clogit(I(status == 'case') ~ driver + suburban + strata(ID), data = backpain)
summary(backpain_glm)
```

## Interpretation

The estimate of the odds ratio of a herniated disc occurring in a driver
relative to a nondriver is 1.93 with a 95% confidence interval of (1.09, 3.44).
Conditional on residence we can say that the risk of a herniated disc occurring
in a driver is about twice that of a nondriver. There is no evidence that where
a person lives affects the risk of lower back pain.


# Tutorial 2

## Prepare workspace and data

Set working directory

```{r}
setwd("E:/Epi_Stat_Matters/LectureNotes2015/Clogit-DrPH-Epid-2015-16")

```

## Read data

```{r}
# source for data 
# use read.table('http://www.medepi.net/data/mi.txt',sep="")
data1<-read.csv('dataclogit.csv',header = TRUE)
```

## Overview of data 

View the first 6 observations

```{r}
head(data1)
```

## Convert and Labels Data

Convert the variables mi, smk and ecg to categorical variables

```{r}
data1$mi2 <- factor(data1$mi, levels = c(1,0), labels = c("Case","Control"))
data1$smk2 <- factor(data1$smk, levels = c(0,1), labels=c("Not current","Current"))
data1$ecg2 <- factor(data1$ecg, levels = c(0,1), labels=c("Normal","Abnormal"))
str(data1,15)
```

## Quick Look at Data

View the first 15 observations

```{r}
head(data1,15)
# load survival package to run clogit
```

Perfom data exploration

```{r expl}
library(psych)
describe(data1)
```

Now, by groups

```{r bygrps}
describeBy(data1, group = 'mi2')
```

# Run clogit Function

This requires **survival** package

```{r}
library("survival")
res.clog <- clogit(I(mi2=='Case') ~ smk2 + strata(match), data = data1)
summary(res.clog)
```

## Reference

Can read here <http://www.medepi.net/docs/ph251d2013fall_REGRESSION-CHAP.pdf>

# Tutorial 3

## Dataset 2

Let us play with another dataset. This tutorial comes from: 

<https://denishaine.wordpress.com/2013/03/22/veterinary-epidemiologic-research-glm-part-4-exact-and-conditional-logistic-regressions/>

## Salmonella outbreak dataset in stata format

## Read stata file

load foreign library to read stata file

```{r}
library(foreign)
# read data
data2 <- read.dta('sal_outbrk.dta', convert.factors = T)
# see variable names
names(data2)
```

## Quickly examine data

```{r}
head(data2)
```

## Run the clogit analysis

Load survival package to run analysis. clogit is a function under survival package (survival::clogit)

```{r}
library(survival)
mod7 <- clogit(I(casecontrol=='case') ~ slt_a + strata(match_grp), data = data2)
summary(mod7)
```

# Tutorial 4 

## Dataset 3

This comes from hosmer book

## Reference

This data and tutorial come from example : <http://www.ats.ucla.edu/stat/stata/examples/alr2/alr2stata7.htm>

## Check data

```{r}
readLines('lowbwt11.dat', n=5)
```

## Import data

We will read a .dat data. 

```{r}
data3<-read.table('lowbwt11.dat')
```

## Quickly view data

Overview of data

```{r}
head(data3,10)
```

## Names the columns

We give names to columns 

```{r}
colnames(data3)<-c('pair','low','age','lwt','race','smoke','ptd','ht','ui')
```

## Declare variables as factors (categorical variables)

Using lapply is fast

```{r}
data3[,c(2,5:9)]<-lapply(data3[,c(2,5:9)],  as.factor)
head(data3)
```

## Specify the levels of the categorical variables

```{r}
levels(data3$low) <- c('bwt>2500g','bwt=<2500g')
levels(data3$race) <- c('white','black','other')
levels(data3$smoke) <- c('no','yes')
levels(data3$ptd) <- c('none','yes')
levels(data3$ht) <- c('no','yes')
levels(data3$ui) <- c('no','yes')
str(data3)
```

## Perform clogit function

covariate = lwt

```{r lwt}
c.data3 <- clogit(I(low=='bwt=<2500g') ~ lwt + strata(pair), data = data3)
summary(c.data3)
```

```{r smoke}
c.data3sm <- clogit(I(low=='bwt=<2500g') ~ smoke + strata(pair), data = data3)
summary(c.data3sm)
```

# Other issues to consider in clogit

## Test Functional Form for Numerical Variable 

unable to do with mfp with surv

Can refer here <http://www.ats.ucla.edu/stat/stata/examples/alr2/alr2stata7.htm>


## **cut** function to break numerical variables


```{r}
data3$cat.lwt <- cut(data3$lwt, breaks = c(min(data3$lwt)-1, 106.5, 120.0, 136.5, max(data3$lwt)))
table(data3$cat.lwt)
```

Run clogit again

```{r multicovar}
c.data3des <- clogit(I(low=='bwt=<2500g') ~ cat.lwt + smoke + ptd + ht + ui + strata(pair), data = data3)
summary(c.data3des)
```

## Tidy your R output

Nice outputs

```{r}
library(broom)
c.data3des2 <- tidy(c.data3des)
c.data3des2
```

Now, we plot the mid-points to see the pattern of 'linearity in logits'

```{r midpts}
# plot (midpoint vs beta)
plot(c(93.3, 113.5, 128.5, 188.5),c(0, c.data3des2[1:3,2]), type = 'l')
```

## Prediction




```{r}
data3final <- clogit(I(low=='bwt=<2500g') ~ lwt + smoke + ptd + ht + ui + strata(pair), data = data3)
summary(data3final)
nice.op <- tidy(data3final)
write.csv(nice.op,'tableclogit.csv')
```

calculate the probability of a positive outcome conditional on one positive outcome within group
in stata we use 

1.  [ predict probposOC, pc1 ] for probability and 
2.  [ predict LinPred, xb ] for linear predictor (log odds) 

Predict (not as good as stata):
* type = 'expected' gives the predicted probability -  calculates the probability of a positive outcome conditional on one positive outcome within group (strata)  


```{r prdt}
# predicted probability 
data3finalfitted <- predict(data3final, type = 'expected')
cbind(data3[1:10, c(1:3, 4,6:9)], data3finalfitted[1:10])
```

* in a conditional logistic the "expected number of events" is just
exp(eta)/(1 + exp(eta)) where eta is the linear predictor. In stata this is known as the probability of a positive outcome, assuming that the fixed effect is zero. See <http://grokbase.com/t/r/r-help/146gcqqxse/r-prediction-based-on-conditional-logistic-regression-clogit>. Also see below 

```{r }
odds_low <- predict(data3final, type = "risk")
(odds_low/(odds_low+1))[1:10]
```



# References

1. <https://cran.r-project.org/web/packages/HSAUR2/vignettes/Ch_logistic_regression_glm.pdf>
2. <http://grokbase.com/t/r/r-help/146gcqqxse/r-prediction-based-on-conditional-logistic-regression-clogit>
3.  <http://stackoverflow.com/questions/35329585/how-to-get-fitted-values-from-clogit-model>
