---
title: "Analysis of Matched Case-Control Studies"
author: "Kamarul Imran Musa"
date: "14 February 2016"
output: 
  pdf_document:
    toc: true
---

# Matched Case-Control Studies Require Special Logistic Regression Approach

To analyze case-control data with pairwise matching we use conditional logistic 
regression. Cases are individually matched to 1 (or up to 4) control subjects based on matching criteria. These matching criteria represent the confounders to be controlled. Each case and matched control(s) are analyzed as separate strata. 
For example, if there are 150 cases, then there are 150 strata that need to be fitted. This is accomplished with the Cox proportional hazards regression functions available in the Survival package.

# Motivation

With matched pairs data the form of the logistic model involves the probability, $\phi$, that in matched pair number i, for a given value of the explanatory
variable the member of the pair is a case. Specifically the model is

$$logit(\phi_{i}) = \alpha_{i} + \beta x$$
The odds that a subject with x = 1 is a case equals $exp(\beta)$ times the odds
that a subject with x = 0 is a case.

# Tutorial 1

```{r}
library(HSAUR2)
head(backpain)
```

Describe data

```{r}
library(psych)
describe(backpain)
```

Perform survival::clogit

```{r}
library(survival)
backpain_glm <- clogit(I(status == 'case') ~ driver + suburban + strata(ID), data = backpain)
summary(backpain_glm)
```

Interpretation

The estimate of the odds ratio of a herniated disc occurring in a driver
relative to a nondriver is 1.93 with a 95% confidence interval of (1.09, 3.44).
Conditional on residence we can say that the risk of a herniated disc occurring
in a driver is about twice that of a nondriver. There is no evidence that where
a person lives affects the risk of lower back pain.


# Tutorial 2

# Prepare workspace and data

## Set working directory

```{r}
setwd("E:/Epi_Stat_Matters/LectureNotes2015/Clogit-DrPH-Epid-2015-16")

```

## Read data

```{r}
# data1<-read.table('http://www.medepi.net/data/mi.txt',sep="")
# write.csv(data1,'dataclogit.csv')
data1<-read.csv('dataclogit.csv',header = TRUE)
```

## Give variables names

```{r}
#colnames(data1)<-c("x","match","person","mi","smk","sbp","ecg")
head(data1)
summary(data1)
```

# Convert and Labels Data

```{r}
data1$mi <- factor(data1$mi, levels=0:1, labels=c("No","Yes"))
data1$smk <- factor(data1$smk, levels=0:1, labels=c("Not current","Current"))
data1$ecg <- factor(data1$ecg, levels=0:1, labels=c("Normal","Abnormal"))
str(data1)
head(data1,15)
```

# Quick Look at Data

```{r}
str(data1)
head(data1,15)
# load survival package to run clogit
```

# Run clogit Function

This requires **survival** package

```{r}
library("survival")
summary(data1)
res.clog<-clogit(I(mi=='Yes')~smk+strata(match),data=data1)
summary(res.clog)
```

## Reference

Can read here <http://www.medepi.net/docs/ph251d2013fall_REGRESSION-CHAP.pdf>

# Dataset 2

Let us play with another dataset. This tutorial comes from: 

<https://denishaine.wordpress.com/2013/03/22/veterinary-epidemiologic-research-glm-part-4-exact-and-conditional-logistic-regressions/>

## Salmonella outbreak dataset in stata format

## Read stata file

load foreign library to read stata file

```{r}
library(foreign)
# read data
data2<-read.dta('sal_outbrk.dta',convert.factors = T)
# see variable names
names(data2)
# quick view of the first 10 rows
```

## Quickly examine data

```{r}
head(data2,10)
# load survival package to run clogit
```

## Load survival package ti run analysis

```{r}
library(survival)
list.files()
mod7 <- clogit(I(casecontrol=='case') ~ slt_a + strata(match_grp), data = data2)
summary(mod7)
```

# Dataset 3

This comes from hosmer book

## Reference

This data and tutorial come from example : <http://www.ats.ucla.edu/stat/stata/examples/alr2/alr2stata7.htm>

## check data

```{r}
readLines('lowbwt11.dat', n=5)
```

## Import data

```{r}
data3<-read.table('lowbwt11.dat')
```

## Quickly examine data

```{r}
head(data3,10)
```

## Names the columns

```{r}
colnames(data3)<-c('pair','low','age','lwt','race','smoke','ptd','ht','ui')
```

## Declare variables as factors (categorical variables)

```{r}
data3[,c(2,5:9)]<-lapply(data3[,c(2,5:9)], as.factor)
```

## Specify the levels of the categorical variables

```{r}
levels(data3$low)<-c('bwt>2500g','bwt=<2500g')
levels(data3$race)<-c('white','black','other')
levels(data3$smoke)<-c('no','yes')
levels(data3$ptd)<-c('none','yes')
levels(data3$ht)<-c('no','yes')
levels(data3$ui)<-c('no','yes')
str(data3)
```

## Perform clogit function


```{r}
c.data3<-clogit(I(low=='bwt=<2500g')~lwt+strata(pair),data=data3)
summary(c.data3)
c.data3sm<-clogit(I(low=='bwt=<2500g')~smoke+strata(pair),data=data3)
summary(c.data3sm)
```

# Other matters

## Test Functional Form for Numerical Variable 

unable to do with mfp with surv

Can refer here <http://www.ats.ucla.edu/stat/stata/examples/alr2/alr2stata7.htm>


## **cut** function to break numerical variables

but before for min 


```{r}
data3$cat.lwt<-cut(data3$lwt, breaks=c(min(data3$lwt)-1,106.5,120.0,136.5,max(data3$lwt)))
table(data3$cat.lwt)
c.data3des<-clogit(I(low=='bwt=<2500g')~cat.lwt+smoke+ptd+ht+ui+strata(pair),data=data3)
summary(c.data3des)
```

## Tidy your R output

Nice outputs

```{r}
library(broom)
c.data3des2<-tidy(c.data3des)
# plot (midpoint vs beta)
plot(c(93.3,113.5,128.5,188.5),c(0,c.data3des2[1:3,2]),type = 'l')
```

## Prediction

Predict (not as good as stata)


```{r}
data3final<-clogit(I(low=='bwt=<2500g')~lwt+smoke+ptd+ht+ui+strata(pair),data=data3)
summary(data3final)
nice.op<-tidy(data3final)
write.csv(nice.op,'tableclogit.csv')

#calculate the probability of a positive outcome conditional on one positive outcome within group
# in stata [ predict probposOC, pc1 ] for probability and [ predict LinPred, xb ] for linear 
# predictor
data3finalfitted<-predict(data3final,type= 'expected')
#to predict linear prediction, not sure in R
# need to find out how 'risk' works for below?????
data3finalfitted<-predict(data3final,type= 'risk')
data3finalfitted<-predict(data3final,type= 'terms')
data3finalfitted<-predict(data3final,type= 'lp')



```

# References

1. <https://cran.r-project.org/web/packages/HSAUR2/vignettes/Ch_logistic_regression_glm.pdf>
